---
title: "Lab 09"
author: 'Group 2-8: Dav King, Vivian Zhang, Reesey Du Pont, Eesha Yaqub'
subtitle: 'Due April 15 at 11:59 PM'
output:
  pdf_document: default
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: console
---

```{r include = FALSE}
knitr::opts_chunk$set(warning = FALSE, 
                      message = FALSE)
```


```{r load-packages}
library(tidyverse)
library(tidymodels)
library(knitr)
library(viridis)
```

```{r load-data}
parkinsons = read_csv("parkinsons.csv")
```


# Exercise 1

```{r rm-underscore}
park = parkinsons %>%
  mutate(name = str_remove_all(name, "_[^_]+$"))
```

```{r ex-1}
park %>% 
  filter(status == 0) %>%
  select(name) %>% 
  unique() %>% 
  kable()
```

# Exercise 2

```{r ex-2}
test <- park %>% 
  group_by(status) %>% 
  arrange(name) %>% 
  slice(1:24) %>% 
  ungroup()
nrow(test)
train <- park %>% 
  group_by(status) %>% 
  arrange(name) %>% 
  slice(-c(1:24)) %>% 
  ungroup()
nrow(train)
```

# Exercise 3

```{r ex-3}
park %>%
  mutate(PD_status = ifelse(status == 0, "healthy", "PD")) %>%
  ggplot(mapping = aes(x = HNR, y = shimmer, color = PD_status)) + 
  geom_point() + 
  labs(x = "HNR", y = "Variation in Amplitude (dB)", 
       title = "Relationship Between HNR and Variation in Amplitude (dB)") + 
  scale_color_viridis(discrete = TRUE, option = "D", name = "PD Status")
```

We observe a non-linear, negative correlation overall between HNR and 
amplitude variance. More specifically, those with PD had a larger range in 
amplitude variance, while the amplitude variance of healthy individuals was more 
clustered at lower values. Similarly, no healthy participant had an HNR lower 
than 17.5, while participants with PD had HNR values getting even lower than 10.

# Exercise 4
```{r ex-4}
fit_1 <- logistic_reg() %>%
  set_engine("glm") %>%
  fit(as.factor(status) ~ HNR + PPE + jitter + shimmer,
      data = train, family = "binomial")
tidy(fit_1)
```

There are three predictors that are significant at the alpha = 0.05 level: 
HNR, PPE, and shimmer. All three of these predictors are significant because
their p-value levels are less than alpha = 0.05. 

# Exercise 5

```{r ex-5}
prediction = predict(fit_1, test, type = "prob")
test_result = test %>% 
  mutate(predicted_prob_pd = prediction$.pred_1) %>% 
  mutate(prob_50 = if_else(predicted_prob_pd > 0.5, 1, 0)) %>% 
  mutate(prob_75 = if_else(predicted_prob_pd > .75, 1, 0)) %>% 
  mutate(prob_90 = if_else(predicted_prob_pd > .9, 1, 0))
test_result %>% 
  count(status, prob_50)
test_result %>% 
  count(status, prob_75)
test_result %>% 
  count(status, prob_90)
```

With a decision boundary of 50%, our model yields 14 false positives and 0 false
negatives. With a decision boundary of 75%, our model yields 2 false positives 
and 1 false negative. With a decision boundary of 90%, our model yields 0 false 
positives and 7 false negatives.

Using this model as a diagnostic tool for PD, we would want to use the 75% 
boundary. Though it was the only level to have both false positives and false 
negatives, it also had the fewest incorrect measures overall - thus, it is the 
most accurate threshold.
