---
title: "Lab 04"
subtitle: "due [date] 11:59 PM "
author: "Group 2-8: Dav King, Eesha Yaqub, Vivian Zhang, Reesey Du Pont"
editor_options: 
  chunk_output_type: console
output:
  pdf_document: default
---

```{r load-packages, message = FALSE, warning = FALSE}
library(tidyverse)
library(sf)
``` 

```{r load-data, message = FALSE}
states <- st_read("cb_2020_us_state_20m.shp", quiet = TRUE)
census <- read_csv("census_2020_lower48.csv")
head(states)
head(census)
```

# Exercise 1

Be respectful of everyone's opinions. Attend group meetings. Complete all of the
work you're supposed to get through.

# Exercise 2
```{r create_data}
census_data <- states %>%
  full_join(census, by = c("STUSPS" = "state")) %>%
  filter(STUSPS != "AK" & STUSPS != "HI" & STUSPS != "PR")
```

Use the proper join to create the `census_data` data set and please remember to name your chunk.

# Exercise 3

```{r us-plot-1}
ggplot(census_data) +
  geom_sf(aes(fill = pop_2020)) +
  labs(title = "2020 population data by state",
       subtitle ="sorted by raw population", fill = "2020 population" ) +
  theme_bw() +
  scale_fill_gradient2()
```

# Exercise 4

```{r new_var}
census_data <- census_data %>%
  mutate(pop_change = (pop_2020 - pop_2010)/pop_2010)
```

# Exercise 5

```{r us-plot-2}
ggplot(census_data) +
  geom_sf(aes(fill = pop_change)) + 
  scale_fill_gradient2(low = "red", high = "green", midpoint = 0) + 
  labs(title = "Change in Population from 2010 to 2020",
       fill = "Percent Population Change")
```

West Virginia had a net loss of population. The West seems to have generally 
experienced fairly significant population gain, while the area around the 
Mississippi River and the Northeast seems to have remained fairly stagnant 
in terms of population change. 

# Exercise 6

```{r us-plot-3}
census_data %>%
  mutate(seat_change_2010 = seats_2020 - seats_2010) %>%
ggplot(.) +
  geom_sf(aes(fill = seat_change_2010 )) +
  scale_fill_gradient2(low = "red", high = "green", midpoint = 0) + 
  labs(title = "Seat change by state", subtitle = "from 2010 to 2020",
       fill = "Seat change ")
```

The majority of states didn't experience a seat change. A cluster of states in the midwest experienced a loss in seats, and also California. 6 states experienced an increase in seat numbers. 


# Exercise 7

```{r us-plot-4}
census_data %>%
  mutate(seat_change_1910 = seats_2020 - seats_1910) %>%
ggplot(.) +
  geom_sf(aes(fill = seat_change_1910)) +
  scale_fill_gradient2(low = "red", high = "green", midpoint = 0) +
  labs(title = "Change in Apportioned Seats by State",
       subtitle = "From 1910 to 2020", fill = "Change in Seats") +
  theme_bw()
```

Given that the geographic distribution of the US has changed drastically over 
the past 110 years but that the number of Representatives in the House has 
remained fixed over that same time period, the trend in seats clearly reflects 
population shift patterns. In the Northeast and Midwest (and to an extent, parts
of the Deep South), seats were lost - especially in former colonial powerhouses 
Pennsylvania and New York, which are both still large but less so in terms of 
relative size compared to where they were 100 years ago. In much of the West, 
states gained a lot of seats - with the largest gains in California, Texas, and 
Florida (which isn't the West, admittedly).