---
title: "Lab 03"
subtitle: "due January 28th by 11:59 PM "
author: "Dav King"
date: "1/24/2022"
editor_options: 
  chunk_output_type: console
output:
  pdf_document: default
---

```{r load-packages, message = FALSE}
library(tidyverse)
library(viridis)
```

# Exercise 1

```{r largest_population}
midwest %>%
  group_by(state) %>%
  summarize(sumpop = sum(poptotal)) %>%
  arrange(sumpop)
```

Of these five states, Illinois has the largest population at 11,430,602.

# Exercise 2

```{r wisconsin_counties}
midwest %>%
  filter(state == "WI") %>%
  arrange(desc(poptotal)) %>%
  slice(1:3) %>%
  select(county, poptotal)
```

The three most populated counties in Wisconsin are Milwaukee (pop. 959275), Dane
(pop. 367085), and Waukesha (pop. 304715).

# Exercise 3

```{r population_density}
midwest %>%
  mutate(metro = if_else(inmetro == 0, "Not in Metro", "In Metro")) %>%
  group_by(metro) %>%
  summarize(avg_pop_density = mean(popdensity),
            number = n())
```

The mean population density of counties in a metro area is 7205 people per unit 
area, while those not in a metro area have an average population density of 951 
ppl/unit area. There are 150 counties in metro areas and 287 counties not in 
metro areas.

# Exercise 4

```{r college_degree}
midwest %>%
  arrange(desc(percollege)) %>%
  slice(1:5) %>%
  select(county, state, percollege)
```

Washtenaw, MI, is home to Ann Arbor and the University of Michigan; Dane, 
Wisconsin is home to Madison and the University of Wisconsin; Champaign, 
Illinois, is home to the University of Illinois. All of these are flagship 
universities for their states, requiring faculty to be college educated and 
focused on raising students to have the same.

# Exercise 5

```{r common_county_names}
midwest %>%
  group_by(county) %>%
  filter(n() == 5) %>%
  summarize(number = n())
```

The counties that occur in all five states are Crawford, Jackson, and Monroe.

# Exercise 6

```{r midwestern_diversity}
midwest %>%
  mutate(nonwhite = if_else(percwhite > 90, 
                        "Less than 10% Nonwhite", "At Least 10% Nonwhite")) %>%
ggplot(aes(x = state, fill = forcats::fct_rev(nonwhite))) +
  geom_bar(position = "fill") +
  scale_fill_viridis(discrete = TRUE, option = "D") +
  labs(title = "How Diverse Are Midwestern States?",
       x = "State", y = "Percent of Counties More Than 10% Nonwhite",
       fill = "Percent Nonwhite") +
  theme_bw()
```

In general, midwestern states are not very diverse at all. No states are over 
25% counties which are at least 10% nonwhite, with the greatest proportion being
Michigan at around 20% and then Illinois at maybe 15%. Then, it is Ohio, 
Wisconsin, and in very last Indiana containing the smallest proportions of 
diverse counties in the midwest.