---
title: "Homework #02: Data Wrangling and Joins"
subtitle: "due [date] 11:59 PM "
author: "Dav King"
date: "1/28/22"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

# Load Packages and Data

```{r load-packages, message = FALSE, warning = FALSE}
library(tidyverse)
library(viridis)
``` 

```{r read-in-data, message = FALSE, warning = FALSE}
natunivs <- read_csv("NatUnivs.csv")
slacs <- read_csv("SLACs.csv")
presvote_pop <- read_csv("PresVote_Population.csv")
```

# Exercise 1

```{r create_data}
full_data <- natunivs %>%
  full_join(slacs, by = c("school", "state",
                             "rank_2022", "rank_2021", "natuniv_slac")) %>%
  inner_join(presvote_pop, by = c("state" = "abbrev"))
```

# Exercise 2

```{r most_schools}
full_data %>%
  group_by(state) %>%
  summarize(nSchools = n()) %>%
  arrange(desc(nSchools)) %>%
  slice(1:5) %>%
  select(state)
```
The 5 states which have the most schools in the `full_data` data set are 
California, Massachusetts, New York, Pennsylvania, and Ohio, respectively.

# Exercise 3

```{r population}
presvote_pop %>%
  anti_join(full_data, by = c("abbrev" = "state")) %>%
  arrange(desc(`2020pop`)) %>%
  select("abbrev", "2020pop")
```

The state with the largest population that does not have a school in `full_data`
is Arizona.

# Exercise 4

```{r copy_plot}
full_data %>%
  mutate(winner = if_else(bidenvotes > trumpvotes, "Biden", "Trump")) %>%
  ggplot(., aes(x = winner, fill = natuniv_slac)) +
  geom_bar() +
  labs(title = "Political Leanings of States with Top Ranked Schools",
       x = "Who won in 2020?", y = "Number of Schools",
       fill = "Type of Institution") +
  scale_fill_viridis(discrete = TRUE, option = "D")
```

In general, the states that Biden won contained far more top ranked schools
(over 80) than the states Trump won (roughly 25). In general, the proportion 
overall seems to be about 50/50 between National Liberal Arts Colleges and 
National Universities, with no notable difference in their distribution in Trump
versus Biden states.

# Exercise 5

```{r school_pop_plot, message = F}
counts <- full_data %>%
  group_by(state) %>% 
  mutate(count = n())
full_data %>%
  full_join(y=counts, by = c("school")) %>%
  ggplot(., aes(x = `2020pop.x`, y = count)) +
  geom_point() +
  geom_smooth(method = lm) +
  labs(title = "Population of US States and Number of Top-Ranked Schools",
       x = "2020 Population", y = "Number of Top Ranked Schools") +
  theme_bw()
```
This graph, using a linear regression line, shows a clear positive relationship
between 2020 population and number of top ranked schools in a state. However,
this appears to be something of an overplotting error. The data are not very 
linearly distributed, and the linear regression line owes most of its strength
to a few outliers than it does a true underlying data structure (which there 
may well be in a relationship with a discrete variable like a number of 
schools). Consider, for example, the graph below - fitted with a Loess 
regression line instead of a linear one. It suggests that the relationship is
almost cubic - certainly, not a standard linear relationship.

```{r loess_graph}
full_data %>%
  full_join(y=counts, by = c("school")) %>%
  ggplot(., aes(x = `2020pop.x`, y = count)) +
  geom_point() +
  geom_smooth(method = loess) +
  labs(title = "Population of US States and Number of Top-Ranked Schools",
       subtitle = "With Loess Regression Line",
       x = "2020 Population", y = "Number of Top Ranked Schools") +
  theme_bw()
```

# Exercise 6

```{r nc_schools}
full_data %>%
  filter(state == "NC") %>% 
  group_by(school) %>% 
  mutate(rank_diff = rank_2021 - rank_2022) %>%
  select(school, rank_diff)
```
There are four schools in North Carolina in this dataset of top schools - Duke
University, Davidson College, Wake Forest University, and our neighbors eight
miles down the road. The "University" of North Carolina and Wake Forest did not
change their rankings from 2021 to 2022, while Davidson improved 2 places and 
Duke increased by 3 (the Duke Difference).

# Exercise 7

```{r political_pop}
full_data %>%
  mutate(biden_share = bidenvotes / (bidenvotes + trumpvotes)) %>%
  group_by(natuniv_slac) %>% 
  summarize(mean_Biden = mean(biden_share), mean_pop = mean(`2020pop`))
```

There seem to be no real differences in Biden vote share between the two groups
of colleges - 56.3% of the vote in one and 57.2% of the vote in the other. 
However, there is a notable difference in how populous th states containing the
two types of schools are. States with National Liberal Arts Colleges tend to 
average around 14 million people, while states with National Universities tend
to average around 16 million.