---
title: 'Homework #03: The Joy of Probability'
author: "Dav King"
date: "2/17/2022"
output:
  pdf_document: default
  html_document:
    df_print: paged
subtitle: 'due February 24 11:59 PM '
editor_options:
  chunk_output_type: console
---

# Load Packages and Data

```{r load-packages, message = FALSE, warning = FALSE}
library(tidyverse)
library(fivethirtyeight)
library(viridis)
library(knitr)
``` 

# Exercise 1

```{r ex-1}
bob_ross %>% 
  count(tree) %>% 
  mutate(tree_prob = n/sum(n))
```

There are 403 episodes, in 361 of which a tree was painted. The probability that
a randomly selected episode featured a tree is 0.896.

# Exercise 2

```{r ex-2}
bob_ross %>% 
  filter(guest == 1) %>% 
  summarize(prob_steve = mean(steve_ross))


bob_ross %>% 
  filter(guest == 1) %>% 
  mutate(steve = if_else(steve_ross == 1, "Steve!", "Not Steve :(")) %>% 
  mutate(mount = if_else(mountain == 1, "Yes!", "No :(")) %>% 
  ggplot(aes(x = steve, fill = factor(mount))) +
  geom_bar(position = "fill") +
  labs(title = "Did Steve Ross Like Mountains More Than Other Guests?",
       x = "Guest on the Show", y = "Mountains Painted",
       fill = "Was a Mountain Painted?") +
  scale_fill_viridis(discrete = T) +
  theme_bw()
```
The conditional probability of Steve Ross being the guest on the show is 0.5. He
did indeed like to paint mountains more than other guests - or at least, he did 
so with far greater frequency.

# Exercise 3

```{r ex-3}
ross_paintings <- bob_ross %>% 
  filter(guest == 0)
```

# Exercise 4

```{r ex-4}
ross_paintings %>% 
  filter(cirrus == 1 & cumulus == 1) %>% 
  select(episode, cirrus, cumulus)
```

They are not disjoint - though there is only one instance in which the two 
clouds were painted together, there was still such an instance; thus, they 
cannot be mutually exclusive.

# Exercise 5

```{r ex-5}
M <- ross_paintings %>% 
  filter(cabin == 1) %>% 
  count(cabin)

X <- ross_paintings %>% 
  filter(cabin == 1) %>% 
  filter(lake == 1) %>% 
  count(lake)

set.seed(2182022) # don't change the seed
num_lakes = rbinom(100000, M$n, prob = 0.5)
cabin_lakes = data.frame(num_lakes)

cabin_lakes %>% 
  mutate(prob_lakes = if_else(num_lakes <= X$n, 1, 0)) %>% 
  summarize(answer = mean(prob_lakes))
```

The probability that Bob Ross painted `X` or fewer lakes, given that he flipped
a fair coin to decide whether or not to paint one every time, is 0.00999. 68 
paintings feature a cabin, and 24 of those also feature a lake.

# Exercise 6

```{r ex-6}
A <- ross_paintings %>% 
  summarize(prob_mountain = mean(mountain), prob_river = mean(river))

B <- ross_paintings %>% 
  filter(river == 1) %>% 
  summarize(mountain_given_river = mean(mountain))

(B$mountain_given_river * A$prob_river)/A$prob_mountain

A$prob_mountain
B$mountain_given_river
```

Bob Ross does not paint mountains independent of whether or not he paints 
rivers. The probability that he paints a mountain is 0.391. If they were 
independent, we would expect the probability of a mountain to be painted given 
that a river was painted to also be 0.391. However, in paintings where a river 
was painted, this probability is 0.387. Though this difference is slight, it is
still present; thus, we can safely say that these two events are not 
independent.

# Exercise 7

Does the relative prevalence of different types of water in this show's 
paintings vary according to the presence of guests on the show and/or an
episode's place within a season?

```{r ex-07-1}
ross3 <- bob_ross %>% 
  mutate(water_type = case_when(
    waves == 1 & ocean == 0 & waterfall == 0 & river == 0 &
      lake == 0 ~ "Waves",
    waves == 0 & ocean == 1 & waterfall == 0 & river == 0 &
      lake == 0 ~ "Ocean",
    waves == 0 & ocean == 0 & waterfall == 1 & river == 0 &
      lake == 0 ~ "Waterfall",
    waves == 0 & ocean == 0 & waterfall == 0 & river == 1 &
      lake == 0 ~ "River",
    waves == 0 & ocean == 0 & waterfall == 0 & river == 0 &
      lake == 1 ~ "Lake",
    T ~ "Two or More"
  )) %>% 
  mutate(nom_guest = if_else(guest == 0, "No Guest", "Guest"))

ross_palette <- c("River" = "#4E1500", "Ocean" = "#FFEC00",
                  "Waterfall" = "#0A3410", "Two or More" = "#021E44",
                  "Lake" = "#C79B00")

ross3 %>% 
  group_by(water_type, nom_guest) %>% 
  summarize(water_presence = n()) %>% 
  ggplot(aes(x = reorder(water_type, -water_presence),
                  y = water_presence, fill = water_type)) +
  geom_col() +
  labs(title = "Prevalence of Different Types of Water in Bob Ross Paintings",
       x = "Water Type", y = "Number of Episodes") +
  theme_bw() +
  coord_flip() +
  scale_fill_manual(values = ross_palette) +
  guides(fill = "none") +
  facet_wrap(~ nom_guest, scales = "free_x")
```

```{r ex-07-2}
pop <- count(ross3)

pWater <- function(water){
  ross3 %>% 
    filter(water_type == water) %>% 
    count(water_type) %>% 
    mutate(p = n/pop)
}

pWaterfall <- pWater("Waterfall")
pOcean <- pWater("Ocean")
pRiver <- pWater("River")
pLake <- pWater("Lake")
pTwo <- pWater("Two or More")

marginal_p_water <- c(pLake$p$n, pOcean$p$n, pRiver$p$n,
                      pTwo$p$n, pWaterfall$p$n)

ross4 <- ross3 %>% 
  count(water_type, nom_guest) %>% 
  pivot_wider(names_from = nom_guest, values_from = n)

ross4$marginal_p_water <- marginal_p_water
kable(ross4)
```

This graph and table suggest that there is likely a difference in the prevalence
of each type of water depending on the presence of a guest, but does not 
necessarily statistically confirm it - this, however, will.

```{r ex-07-3}
given <- function(g, water){
  ross3 %>% 
    filter(nom_guest == g) %>% 
    count(water_type) %>% 
    mutate(p = n/sum(n)) %>% 
    filter(water_type == water)
}

p_guest_ocean <- given("Guest", "Ocean")
p_guest_river <- given("Guest", "River")
p_guest_lake <- given("Guest", "Lake")
p_guest_two <- given("Guest", "Two or More")

guest_probs <- c("X", p_guest_ocean$p, p_guest_river$p,
                 p_guest_lake$p, p_guest_two$p)

p_no_guest_wf <- given("No Guest", "Waterfall")
p_no_guest_ocean <- given("No Guest", "Ocean")
p_no_guest_river <- given("No Guest", "River")
p_no_guest_lake <- given("No Guest", "Lake")
p_no_guest_two <- given("No Guest", "Two or More")

no_guest_probs <- c(p_no_guest_wf$p, p_no_guest_ocean$p, p_no_guest_river$p,
                    p_no_guest_lake$p, p_no_guest_two$p)

A <- "X"
B <- p_guest_ocean$p == p_no_guest_ocean$p
C <- p_guest_river$p == p_no_guest_river$p
D <- p_guest_lake$p == p_no_guest_lake$p
E <- p_guest_two$p == p_no_guest_two$p

across_conditions <- c(A, B, C, D, E)

A1 <- "X"
B1 <- p_guest_ocean$p == pOcean$p$n
C1 <- p_guest_river$p == pRiver$p$n
D1 <- p_guest_lake$p == pLake$p$n
E1 <- p_guest_two$p == pTwo$p$n

guest_vs_baseline <- c(A1, B1, C1, D1, E1)

A2 <- p_no_guest_wf$p == pWaterfall$p$n
B2 <- p_no_guest_ocean$p == pOcean$p$n
C2 <- p_no_guest_river$p == pRiver$p$n
D2 <- p_no_guest_lake$p == pLake$p$n
E2 <- p_no_guest_two$p == pTwo$p$n

no_guest_vs_baseline <- c(A2, B2, C2, D2, E2)

cProb_table <- data.frame(ross4$water_type, marginal_p_water, guest_probs,
                          no_guest_probs, guest_vs_baseline,
                          no_guest_vs_baseline, across_conditions)
cProb_table %>% 
  rename(
    "Water Type" = ross4.water_type,
    "Marginal Probability" = marginal_p_water,
    "Probability Given Guest" = guest_probs,
    "Probability Given No Guest" = no_guest_probs,
    "Guest vs Baseline" = guest_vs_baseline,
    "No Guest vs Baseline" = no_guest_vs_baseline,
    "Guest vs No Guest" = across_conditions) %>% 
  kable()
```

As we can see, in none of these conditions (for any type of water, guest or no
guest as compared both to the total dataset and to each other) are the marginal 
probabilities equal to one another; thus, these two events are not independent 
of one another.

```{r ex-07-4}
ross3 %>% 
  group_by(water_type, nom_guest) %>% 
  mutate(water_presence = n()) %>% 
  ggplot(aes(x = factor(episode_num),
             fill = reorder(water_type, water_presence))) +
  geom_bar(position = "fill") +
  labs(title = "Relative Prevalence of Different Types of Water Across Season",
       x = "Episode Number", y = "Proportion", fill = "Type of Water") +
  theme_bw() +
  scale_fill_manual(values = ross_palette)
```

```{r ex-07-5}
ross3 %>% 
  group_by(water_type, nom_guest) %>% 
  mutate(water_presence = n()) %>% 
  ggplot(aes(x = factor(episode_num),
             fill = reorder(water_type, water_presence))) +
  geom_bar(position = "fill") +
  labs(title = "Relative Prevalence of Different Types of Water Across Season",
       x = "Episode Number", y = "Proportion", fill = "Type of Water") +
  theme_bw() +
  scale_fill_manual(values = ross_palette) +
  facet_wrap(~ nom_guest)
```

These once again suggest both that A) there is a difference in the relative 
prevalence of each type of water across the season and B) that this varies 
according to whether or not there was a guest on the show; but again, let us 
turn to statistics:

```{r ex-07-6}
timeline <- ross3 %>%
  #For simplicity's sake, we will group the episodes into three chunks
  mutate(chunked = case_when(
    episode_num <= 4 ~ "Early",
    episode_num <= 8 ~ "Middle",
    T ~ "Late"
  )) %>% 
  mutate(chunked = factor(chunked, levels = c("Early", "Middle", "Late"),
                          ordered = T)) %>%
  arrange(chunked)

time_prob <- function(time){
  timeline %>% 
    filter(chunked == time) %>% 
    count(water_type) %>% 
    mutate(p = n/sum(n))
}

early_prob <- time_prob("Early")
middle_prob <- time_prob("Middle")
late_prob <- time_prob("Late")

early_prob
middle_prob
late_prob
#In the interest of time, I won't do the logic calculations - we can all see 
#that none of those numbers are the same across groups
```

```{r ex-07-7}
split_time_prob <- function(time, g){
  timeline %>% 
    filter(chunked == time & guest == g) %>% 
    count(water_type) %>% 
    mutate(p = n/sum(n))
}

early_guest <- split_time_prob("Early", 1)
middle_guest <- split_time_prob("Middle", 1)
late_guest <- split_time_prob("Late", 1)

early_no_guest <- split_time_prob("Early", 0)
middle_no_guest <- split_time_prob("Middle", 0)
late_no_guest <- split_time_prob("Late", 0)

early_prob
early_guest
early_no_guest

middle_prob
middle_guest
middle_no_guest

late_prob
late_guest
late_no_guest
```
The only value held in common (and thus even remotely plausible to be 
independent) anywhere in this calculation is a 50% chance of two or more types 
of water in both the early and middle chunks of the season on shows that had 
guests - clearly, these values cannot be considered independent of one another.