---
title: 'Homework 05: Paint a plot, pick a package'
author: 'Dav King'
date: "April 14, 2022"
output:
  pdf_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(warning = FALSE, 
                      message = FALSE)
```

## Load packages & data

```{r load-packages-data}
library(tidyverse)
library(nflreadr)
library(ggridges)
``` 


## Exercise 1

```{r ex-1, fig.height = 3, fig.width = 6}
iris %>% 
  ggplot(aes(x = Sepal.Length, y = Petal.Length, color = Species)) +
  geom_point(aes(size = Sepal.Width, alpha = Petal.Width)) +
  scale_x_log10() +
  scale_y_time() +
  geom_path(size = 2) +
  geom_dotplot(aes(y = Species)) +
  geom_violin() +
  theme(panel.background = element_rect(fill = "hotpink", color = "hotpink"),
        panel.grid.major = element_line(color = "#BF5700", size = 2),
        panel.grid.minor = element_line(color = "#39FF14"))
```

## Exercise 2

This package is called `nflreadr`, which is hosted on CRAN. The packages enables
people to read nflverse data, essentially giving them access to all of the 
play-by-play data from each recorded NFL season alongside draft picks, contract 
information and much much more. With this package, I was able to download the 
play-by-play information for the complete 2021 NFL season and make a graphic of 
the relative density of each type of play depending on how much time was left in
the half (or overtime). 

```{r ex-2, fig.height = 10}
df <- load_pbp(2021)
df %>% 
  filter(!is.na(play_type)) %>% 
  ggplot(aes(x = half_seconds_remaining, y = play_type, fill = play_type)) +
  geom_density_ridges(alpha = 0.5) +
  scale_x_reverse() +
  facet_wrap(~ game_half, ncol = 1) +
  theme_bw() +
  guides(fill = "none") +
  labs(title = "When do Different NFL Plays Happen?", x = "Seconds Remaining",
       y = "Type of Play") +
  theme(plot.title = element_text(hjust = 0.5))
```

