---
title: "Homework #01: Data Visualization"
subtitle: "due January 27, 2021 11:59 PM "
author: "Dav King"
date: "1/20/22"
output: pdf_document
---

# Load Packages

```{r load-packages, message = FALSE, warning = FALSE}
library(tidyverse) 
library(viridis)
library(ggridges)
library(scales)
``` 

# Load Data
```{r load-data, message = F}
anes <- read_csv("anes2020_subset.csv")
```


# Exercise 1

```{r rows_columns}
glimpse(anes)
```

There are 8,280 rows and 6 columns in the `anes` dataset.

# Exercise 2

```{r ideology_plot, warning = FALSE, message = FALSE}
ggplot(anes, aes(x = ideology, fill = as.factor(ideology))) +
  geom_bar() +
  labs(title = "Respondents Arranged by Ideology",
       x = "Ideology (Liberal to Conservative)",
       y = "Number of Respondents") +
  theme_bw() +
  scale_fill_viridis(discrete = TRUE, option= "H") +
  guides(fill = "none")
```

The most common ideology is 4 (on a scale from 1-7), or the people who identify 
themselves as true moderates. In general, respondents tend to label themselves 
as more moderate than they do extreme (regardless of how true this is in 
reality).

# Exercise 3

```{r ideology_by_location, warning = FALSE}
ggplot(anes, aes(x = ideology, fill = urbanrural)) +
  geom_bar(position = "fill") +
  labs(title = "Where Do Liberals, Moderates, and Conservatives Live?",
       x = "Ideology (Liberal to Conservative",
       y = "Number of Respondents",
       fill = "Location") +
  scale_fill_viridis(discrete = TRUE, option = "D", na.value = "Black") +
  theme_bw()
```

People who are more liberal tend to live in cities, while people who are more 
conservative tend to live in more rural places (each of those locations appears 
directly tied to ideology, with no exceptions, relative to other ideologies). 
However, people who are more moderate are relatively more prevalent in the 
suburbs. Though people in small towns are more or less evenly distributed 
ideologically, people who are more conservative tend to live in small towns more 
often than people who are very liberal. The percentage non-responses is almost 
entirely evenly distributed, though it might be a slight bit higher among 
conservatives than liberals.

# Exercise 4

```{r scientist_feelings, warning = FALSE}
ggplot(anes, aes(x = scientists, fill = as.integer(scientists),
                 group = scientists)) +
  geom_histogram(binwidth = 5) +
  labs(title = "How Do Americans Feel About Scientists?",
       x = "Warmth Towards Scientists",
       y = "Number of Respondents") +
  theme_bw() +
  scale_fill_gradient(low = "Blue", high = "Red") +
  guides(fill = "none")
```

In general, people report feeling very warmly towards scientists - the data are 
heavily skewed left. While unimodal, the peak of the data occurs at 100 - the 
maximum possible score. Additionally, responses seem to be clustered at a few 
specific points (almost regardless of bin width). Though this is not in the 
question itself, we would be ill-advised to draw any conclusions from data this 
heavily skewed - it suggests that perhaps there was a flaw in the methodology 
from which this data set was created.

# Exercise 5

```{r fish_hunt_ideology, warning = FALSE, message = FALSE}
anes %>%
  drop_na(hunt_fish) %>%
  mutate(hunted_fished = ifelse(hunt_fish == 0,
                                "Did Not Hunt or Fish", "Hunted or Fished")) %>%
  ggplot(., aes(x = hunted_fished, y = ideology, color = hunted_fished)) +
  geom_boxplot(lwd = 1) +
  labs(title = "Is Hunting or Fishing Related to Ideology?",
       x = "Did Respondents Hunt or Fish?",
       y = "Ideology (Liberal to Conservative") +
  theme_bw() +
  scale_color_viridis(discrete = TRUE, option = "D")+
  guides(color = "none")
```

```{r ridgline_density_plot, warning = FALSE, message = FALSE}
anes %>%
  drop_na(hunt_fish) %>%
  mutate(hunted_fished = ifelse(hunt_fish == 0,
                                "Did Not Hunt or Fish", "Hunted or Fished")) %>%
  ggplot(., aes(x = ideology, y = hunted_fished, fill = hunted_fished)) +
  geom_density_ridges(alpha = 0.5, color = "black", lwd = 1) +
  labs(title = "Is Hunting or Fishing Related to Ideology?",
       x = "Ideology (Liberal to Conservative)",
       y = "Did Respondents Hunt or Fish?") +
  theme_bw() +
  scale_fill_viridis(discrete = TRUE, option = "D") +
  guides(fill = "none")
```

In the graph of the boxplots, it is easy to see the median ideology of each 
group, which is 4 or moderate for `Did Not Hunt or Fish` and 5 or slightly 
conservative for `Hunted or Fished`. It also shows the rest of the summary 
statistics, which suggest a wide distribution in ideology of people who did not 
hunt or fish but a much narrower distribution concentrated on the conservative 
side of the spectrum. However, there are no outliers, perhaps to be expected 
with such a narrow range of possible values.

In the graph of the ridgelines, we do not get to see summary statistics in the 
same way. However, we do get to see individual data points. This shows the 
higher concentration of people who hunted or fished on the Conservative side of 
the spectrum, and reveals that the median value at 5 is actually a number chosen
by relatively few respondents within that category. It also shows the more 
normal distribution of people who did not hunt or fish, who had surprisingly low
rates of `3` and `5`. Unfortunately, this also shows some of the limitation of 
ridgeline plots - those drapes are entirely meaningless (in fact, they provide 
confusing data points that do not actually exist) and smoothing the curve to 
remove them would also eliminate any of the benefits from the ridgeline plot.

The boxplots give the summary statistics, which is helpful because they provide 
the clearest summary for the data's quartiles and thus make the two groups 
immediately comparable. However, what they do not give that the ridgeline plot 
does is individual data points, which remind us that the median can settle on a 
value that is not particularly common.

# Exercise 6

```{r scientists_vs_education, warning = FALSE, message = FALSE}
ggplot(anes, aes(x = education, y = scientists)) +
  geom_point(aes(color = as.factor(education))) +
  scale_color_viridis(discrete = TRUE, option = "H") +
  guides(color = "none") +
  labs(title = "Is Education Related to Views of Scientists?",
       x = "Education", y = "Warmth Towards Scientists") +
  theme_bw() +
  geom_smooth(method = lm)
```
This is a very poor way to visualize data and isn't helpful in the slightest. 
When making scatterplots, we do so with the intention of seeing whether an 
increase in one variable is related to an increase (or decrease) in another. 
However, while this does the same thing, `Education` is not a continuous 
variable in this dataset. Instead, it is an ordinal categorical variable, which 
therefore presents in column-like structures on a scatterplot and is almost 
unreadable as a result. It also requires more specific labels across the x-axis,
as the discrete points of the variable are functionally meaningless. 
Additionally, since the gaps in years of education or any other continuous 
variable are not consistent from one discrete marker to another in this dataset,
there is no interpretable structure in the data and the regression line is 
misleading at best. Scatterplots with discrete data such as these are not 
useful.

# Exercise 7

```{r education_scientists_hunting, warning = FALSE, message = FALSE}
set.seed(18)
anes2 <- anes %>%
  sample_frac(.10)
facet.labs <- c("Hunted or Fished",
                "Did Not Hunt or Fish",
                "Unknown")
names(facet.labs) <- c("1", "0", NA)
ggplot(na.omit(anes2), aes(x = education, y = scientists,
                           color = as.factor(hunt_fish))) +
  geom_point() +
  scale_color_viridis(discrete = TRUE, option = "D") +
  guides(color = "none") +
  facet_grid(cols = vars(hunt_fish),
             labeller = labeller(hunt_fish = facet.labs)) +
  labs(title = "Is Education Related to Views of Scientists?",
       subtitle = "Faceted by Hunting and/or Fishing in the Past Year",
       x = "Education", y = "Warmth Towards Scientists") +
  theme_bw()+
  geom_smooth(method = lm, se = FALSE)
```

In these graphs, almost the same trend as before comes into play regardless of
facet. Though using `Education`, a discrete variable, makes for very poor
scatterplots, once again a slight linear correlation appears to exist between
education and views of scientists. This is seen in both groups - those who did
and did not hunt or fish in the past year - and while the correlation is the
slightest bit stronger (and begins and ends at slightly higher intercepts) for
the group that did not hunt or fish, this difference would not be significant if
we ran statistical tests. Both groups start out at a high level of warmth
towards scientists, and this increases slightly with education in the sample.