---
title: 'AE 10: Probability '
author: "Dav King"
date: "2/10/2022"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(warning = FALSE, 
                      message = FALSE)
```

## Learning goals 

- Introduce probabilities and how we can use them to understand categorical data 
- Create a contingency table using `pivot_wider()` and `kable()`
- Use a contingency table to explore the relationship between two categorical variables.

## Introduction

```{r load-packages}
library(tidyverse)
library(knitr)
```

```{r load-data}
sta199 <- read_csv("sta199-fa21-year-major.csv")
```

For this Application Exercise, we will look at the year in school and majors for students taking STA 199 in Fall 2021. The data set includes the following variables: 

- `section`: STA 199 section
- `year`: Year in school
- `major_category`: Major / academic interest. 
  - For the purposes of this AE, we'll call this the student's "major". 

## Definitions

- The **probability** of an event tells us how likely an event is to occur, and it can take values from 0 to 1, inclusive. It can be viewed as  
  - the proportion of times the event would occur if it could be observed an infinite number of times.
  - our degree of belief an event will happen.
  
- An **event** is the basic element to which probability is applied, e.g. the result of an observation or experiment. 
  - Example: **A** is the event a student in STA 199 is a sophomore. 

- A **sample space** is the set of all possible outcomes. Each outcome in the sample space is **disjoint** or **mutually exclusive** meaning they can't occur simultaneously. 
   - Example: The sample space for year is {First-year, Sophomore, Junior, Senior}

## Exercise 1

Let's take a look at the majors. Note that we have categorized majors so that each student can only be in one major category. 

- What is the sample space for major? You can use code to identify the sample space. 

```{r major-sample-space}
unique(sta199$major_category)
```

- Let's make a table that includes the majors, the number of students in each, and the associated probabilities. 

```{r majors}
sta199 %>% 
  count(major_category) %>% 
  mutate(prop_major = n/sum(n))
```

- What is the probability a randomly selected STA 199 student is a "pubpol only" major?  

```{r prob-pub-pol}
sta199 %>% 
  count(major_category) %>% 
  mutate(prop_major = n/sum(n)) %>% 
  filter(major_category == "pubpol only")
```

- What is the probability a randomly selected STA 199 student is studying statistics? 

```{r prob-stats} 
sta199 %>% 
  count(major_category) %>% 
  mutate(prop_major = n/sum(n)) %>% 
  filter(major_category == "stat + other major" |
           major_category == "stats only")
```

- What is the probability a randomly selected STA 199 student is not a "pubpol only" major? 

```{r prob-not-pub-pol} 
sta199 %>% 
  count(major_category) %>% 
  mutate(prop_major = n/sum(n)) %>% 
  filter(major_category != "pubpol only") %>% 
  mutate(answer = sum(prop_major)) %>% 
  summarize(answer) %>% 
  slice(1)
```


## Exercise 2

Now let's make at table looking at the relationship between year and major.

```{r count-yr-major}
sta199 %>%
  count(year, major_category)
```

We'll reformat the data into a **contingency table**, a table frequently used to study the association between two categorical variables. In this contingency table, each row will represent a year, each column will represent a major, and each cell is the number of students have a particular combination of year and major. 

To make the contingency table, we will use a new function in `dplry` called `pivot_wider()`. It will take the data frame produced by `count()` that is current in a "long" format and reshape it to be in a "wide" format. 

We will also use the `kable()` function in the `knitr` package to neatly format our new table.

```{r pivot-wider}
sta199 %>% 
  count(year, major_category) %>%
  pivot_wider(id_cols = c(year, major_category),#how we identify unique obs
              names_from = major_category, #how we will name the columns
              values_from = n, #values used for each cell
              values_fill = 0) %>% #how to fill cells with 0 observations 
  kable() # neatly display the results
```
- How many students in STA 199 are first-years and in the "econ only" majors category.
6
- How many students in STA 199 are in the "other" major category? 
98

## Exercise 3

For each of the following exercises:

(1) Calculate the probability using the contingency table above. 

(2) Then write code to check your answer using the  `sta199` data frame and `dplyr` functions. 

- What is the probability a randomly selected STA 199 student is a sophomore? 
.397

```{r prob-sophomore}
sta199 %>% 
  count(year) %>% 
  mutate(prop_year = n/sum(n))
```

- What is the probability that a randomly selected STA 199 student is a "compsci only" major? 
.162

```{r prob-cs}
sta199 %>% 
  count(major_category) %>% 
  mutate(prop_major = n/sum(n))
```


- What is the probability that a randomly selected STA 199 student is a sophomore **or** a "compsci only" major? 
.466

```{r prob-sophomore-cs}
sta199 %>% 
  mutate(soph_cs = case_when(
    year == "Sophomore" | major_category == "compsci only" &
                                        year != "Sophomore" ~ 1,
    T ~ 0
  )) %>% 
  count(soph_cs) %>% 
  mutate(answer = n/sum(n))
```

- What is the probability that a randomly selected STA 199 student is a sophomore **and** a "compsci only" major?
.093

```{r prob-sophomore-cs-2}
sta199 %>% 
  count(year, major_category) %>% 
  mutate(prop_maj_year = n/sum(n)) %>% 
  filter(year == "Sophomore" & major_category == "compsci only")
```

## Resources 

- Notes on `pivot_wider` and `pivot_longer`
  - [Click here](https://rstudio-education.github.io/datascience-box/course-materials/slides/u2-d09-tidying/u2-d09-tidying.html#1) for slides
  - [Click here](https://www.youtube.com/watch?v=x3KM5uxaFdI) for video