---
title: "Lab 05"
author: "Group 2-8: Dav King, Eesha Yaqub, Reesey Du Pont, Vivian Zhang"
subtitle: 'due February 18 at 11:59 PM '
output:
  pdf_document: default
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: console
---

```{r include = FALSE}
knitr::opts_chunk$set(warning = FALSE)
```

```{r load-packages, message = FALSE, warning = FALSE}
library(tidyverse)
library(knitr)
library(viridis)
``` 

```{r load-data, message=FALSE}
courage <- read_csv("courage.csv")
```

# Exercise 1

There are 78 observations in the dataset, with each observation representing one
game that the NC Courage played.

# Exercise 2
```{r ex-2}
seasonal_courage = courage %>% 
  mutate(seasonal_category = case_when(
    game_number <= 9 ~ "early",
    game_number <= 17 ~ "middle",
    game_number <= 26 ~ "late"
  )) %>% 
  mutate(win = if_else(result == "win", 1, 0)) %>% 
  group_by(seasonal_category) %>% 
  summarize(prop_win = mean(win))
seasonal_courage
```

This table shows the conditional probability that the Courage win a soccer game,
given the point in the season that the game was played. The conditional 
probability that the team won a game given that it was early in the season is 
0.593, given that it was late in the season is 0.704, and given that it was the 
middle of the season is 0.75.

# Exercise 3

```{r ex-3}
seasonal_courage %>%
   mutate(seasonal_category =
            factor(seasonal_category,
                   levels = c("early", "middle", "late"),
                   ordered = TRUE)) %>%
  arrange(seasonal_category)
```

# Exercise 4

```{r ex-4}
courage %>%
  mutate(Home_Game = case_when(home_team == "NC" ~ 1, T ~ 0)) %>%
  mutate(win = case_when( result == "win" ~ 1, T ~ 0 )) %>% 
  mutate(pwin = mean(win)) %>%
  mutate(pHome_Game = mean(Home_Game)) %>%
  filter(result == "win") %>%
  mutate(pwin_athome = mean(Home_Game)) %>%
  mutate(pwin_Home_Game = ((pwin_athome)*pwin)/ pHome_Game) %>%
  summarize(pwin, pwin_Home_Game)

```
P(win) = 0.679
Conditional Probability = 0.738 
No because these two values are different. If the two events were independent, 
we would expect the probability of a win to be the exact same as the probability 
of a win given that it was a home game. Instead, the probability of winning at 
home is much higher. This suggests a strong home-field advantage for the Courage
- they win much more frequently at home.

# Exercise 5

```{r ex-5}
courage <- courage %>% 
  mutate(home_courage = if_else(home_team == "NC", "home", "away"))
courage %>%
count(result, home_courage) %>% 
  pivot_wider(id_cols = c(result, home_courage),
              names_from = home_courage, 
              values_from = n, 
              values_fill = 0) %>% 
  kable()
```

P(home) = (42/78) = 0.538

P(tie) = (11/78) = 0.141

P(home|tie) = (6/11) = 0.545

Finding P(tie|home) using Bayes' Theorem: 
P(tie|home) = ((0.545)(0.141))/(0.538) = 0.143

Checking Bayes' Theorem using the contingency table: 
P(tie|home) = (6/42) = 0.143 

The probability of a tie is not independent of the Courage playing at home 
or away because when comparing the conditional probability of a tie occurring 
when the courage plays at home (or P(tie|home)) is not equal to the marginal 
probability of a tie occurring (or P(tie)).

# Exercise 6

```{r ex-6-make-df}
courage <- courage %>%
  mutate(total_pts = home_pts + away_pts) %>%
  mutate(courage_pts = if_else(home_courage == "home", home_pts, away_pts)) %>%
  mutate(opponent_pts = if_else(home_courage == "away", home_pts, away_pts))
```


```{r ex-6-plot}
courage %>%
  ggplot(mapping = aes(x = courage_pts, 
                       y = opponent_pts, color = home_courage)) + 
  geom_point() + 
  scale_color_viridis(discrete = TRUE, option = "D") + 
  labs(title = "Points Scored by NC Courage and Opponents", 
       subtitle = "Faceted by Season", x = "NC Courage Points", 
       y = "Opponent Points", color = "Home or Away Game") +
  geom_jitter(width = 0.1, height = 0.1) + 
  geom_abline(slope = 1, intercept = 0) + 
  facet_wrap(~ season)
```

The graphs show that the majority of the highest-scoring games for NC Courage 
were at home, while most of the games where the opponent outscored NC Courage
(or where the opponent just scored high in general, even if they still scored 
lower than NC Courage) were away games. Additionally, NC Courage seems to have 
won the majority of their games, since in the graphs for each season, more 
points lie to the right of the y = x line rather than to the left (indicating 
an NC Courage win). 