---
title: "Homework #04: Bass Mercury"
subtitle: "due March 3rd 11:59 PM "
author: "Dav King"
date: "2/28/2022"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

# Load Packages and Data

```{r load-packages, message = FALSE, warning = FALSE}
library(tidyverse)
library(tidymodels)
library(viridis)
``` 

```{r load-data, message=FALSE, warning=FALSE}
mercury_bass <- read_csv("mercury.csv")
```


# Exercise 1
$H_0: \mu \leq 0.46$

$H_A: \mu > 0.46$

$H_0$: The average mercury level is 0.46 ppm or less

$H_A$: The average mercury level is more than 0.46 ppm

# Exercise 2

```{r ex-2}
set.seed(2)
null_dist <- mercury_bass %>%
  specify(response = mercury) %>%
  hypothesize(null = "point", mu = 0.46) %>%
  generate(reps = 10000, type = "bootstrap") %>%
  calculate(stat = "mean")
```

# Exercise 3

```{r ex-3}
pop_mean <- mercury_bass %>% 
  summarize(pop_mean = mean(mercury)) %>% 
  pull()

null_dist %>% 
  get_p_value(obs_stat = pop_mean, direction = "greater")

visualize(null_dist) +
  shade_p_value(obs_stat = pop_mean, direction = "greater") +
  labs(title = "Null Distribution and Observed Statistic of Mercury Content",
       x = "Mercury, in PPM", y = "Number of Observations")
```

Assuming $\alpha = 0.05$, no null hypothesis has ever been rejected this hard. 
Even I didn't get rejected this hard in middle school. We have significant 
evidence to say that the true population mean is greater than 0.46 ppm mercury
in the bass from these two rivers.

# Exercise 4

```{r ex-4}
set.seed(4)
mercury_bass %>% 
  specify(response = mercury) %>% 
  generate(reps = 10000, type = "bootstrap") %>% 
  calculate(stat = "mean") %>% 
  summarize(lower = quantile(stat, 0.025),
            upper = quantile(stat, 0.975))
```

Given this sample, we are 95% confident that the true population mean mercury 
content in these two rivers is between 1.08 and 1.31 ppm. This is much more 
consistent with our alternative hypothesis because, just as it hypothesized, it 
puts mu at above 0.46 ppm (and $H_0$ is not even contained on this interval, 
the other way to calculate our p-value).

# Exercise 5

$H_0: \mu W = \mu L$

$H_A: \mu W \neq \mu L$

$H_0$: the average mercury content does not differ between bass caught in the 
two rivers

$H_A$: the average mercury content differs significantly between bass caught in 
the two rivers

```{r ex-5}
mercury_bass <- mercury_bass %>% 
  mutate(riverName = if_else(river == 0, "Lumber", "Waccamaw"))
```

# Exercise 6

```{r ex-6}
set.seed(6)
null_dist_2 <- mercury_bass %>% 
  specify(response = mercury, explanatory = riverName) %>% 
  hypothesize(null = "independence") %>% 
  generate(reps = 10000, type = "permute") %>% 
  calculate(stat = "diff in means")

dif_mean <- mercury_bass %>% 
  group_by(riverName) %>% 
  summarize(merc = mean(mercury)) %>% 
  summarize(dif_mean = merc[riverName == "Lumber"] - 
              merc[riverName == "Waccamaw"]) %>% 
  pull()

visualize(null_dist_2) +
  shade_p_value(obs_stat = dif_mean, direction = "both") +
  labs(title = "Null Distribution of Difference in Mercury Between Rivers", 
       x = "Difference in Means", y = "Count",
       caption = "Note: red area is >= the observed statistic")

null_dist_2 %>% 
  get_p_value(obs_stat = dif_mean, direction = "both")
```

At $\alpha = 0.05$, we fail to reject $H_0$. We do not have significant evidence
to suggest that there are different mean levels of mercury in the fish caught in
each of the two rivers.

# Exercise 7

So we don't HARK and find invalid results that look shiny. We need a hypothesis 
that we didn't curate to become an explanation for our data - because that makes
it entirely free of chance.

```{r ex-7}
null_dist_2 %>% 
  get_p_value(obs_stat = dif_mean, direction = "less")
```

Take this, for example. In our original statistical test, we calculated a 
p-value of 0.089 - significant at $\alpha = 0.1$, but not at $\alpha = 0.05$. 
Additionally, we did not have any reason to believe that one river was notably 
higher than the other - hence, the two-tailed p-test. However, if we (after 
looking at the data) decided to conduct a one-tailed p-test off of a directional
$H_A$, we would find a p-value significant at $\alpha = 0.05$ - but not 
necessarily a legitimate one, as we had no reason to investigate that $H_A$.

# Exercise 8

Does the average length of bass differ between the two rivers?

```{r ex-8}
set.seed(1089336)
diff_length <- mercury_bass %>% 
  group_by(riverName) %>% 
  summarize(mean_length = mean(length)) %>% 
  summarize(diff_length = mean_length[riverName == "Lumber"] - 
              mean_length[riverName == "Waccamaw"]) %>% 
  pull()

null_dist_3 <- mercury_bass %>% 
  specify(response = length, explanatory = riverName) %>% 
  hypothesize(null = "independence") %>% 
  generate(reps = 10000, type = "permute") %>% 
  calculate(stat = "diff in means")

visualize(null_dist_3) +
  shade_p_value(obs_stat = diff_length, direction = "both") +
  labs(title = "Null Distribution of the Difference in Lengths Between Rivers",
       x = "Difference in Mean Length", y = "Count",
       caption = "Note: red area >= observed statistic")

#Note: while I am aware the question didn't ask us to calculate the p-value, I
#would hate to not do so after doing all this work
null_dist_3 %>% 
  get_p_value(obs_stat = diff_length, direction = "both")
```

Based on this visualization (ignoring the p-value, since I wasn't actually 
asked to calculate that), it does not seem like there is much difference between
the length of bass found in the two rivers beyond random chance.